---
title: "Custom Color Palettes"
output:
  github_document:
    toc: true
    toc_depth: 3 
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


> **DISCLAIMER:** This is actually a copy pasted document from these sources. [1] [Creating Corporate Color Palettes for GGPlot2](https://drsimonj.svbtle.com/creating-corporate-colour-palettes-for-ggplot2)


## Creating Corporate Color Palettes for ggplot2

Basic steps are:

- creating a Named vector of hex codes for the colors
- creating a function to access the color codes
- creating named list of color palettes
- creating function to access the color palettes
- creating ggplot2 compatible scale function to use the palettes


```{r}
library(ggplot2)

theme_set(theme_minimal())
```

#### creating the named vector

```{r}
# drsimonj corporate colors

drsimonj_colors <- c(
  `red`        = "#d11141",
  `green`      = "#00b159",
  `blue`       = "#00aedb",
  `orange`     = "#f37735",
  `yellow`     = "#ffc425",
  `light_grey` = "#cccccc",
  `dark_grey`  = "#8c8c8c")
```

#### the accessor funciton

```{r}
drsimonj_cols <- function(...) {
  cols <- c(...)
  
  if (is.null(cols)) {
    return(drsimonj_colors)
  }
  
  drsimonj_colors[cols]
}
```


```{r}
drsimonj_cols()

drsimonj_cols("blue", "red")
```

#### combining color into palettes

```{r}
drsimonj_palettes <- list(
  main = drsimonj_cols("blue", "green", "yellow"),
  cool = drsimonj_cols("blue", "green"),
  hot = drsimonj_cols("yellow", "orange", "red"),
  mixed = drsimonj_cols("blue", "green", "yellow", "orange", "red"),
  grey = drsimonj_cols("light_grey", "dark_grey")
)
```
#### creating palettes accessor and creator function

```{r}
drsimonj_pal <- function(palette = "main", reverse = FALSE, ...) {
  pal <- drsimonj_palettes[[palette]]
  
  if (reverse) pal <- rev(pal)
  
  colorRampPalette(pal, ...)
}
```

`colorRamPalette` will return another function that will enable us create shades between our original colors.


```{r}
drsimonj_pal("cool")

drsimonj_pal("cool")(2)

drsimonj_pal("cool")(10)

```

#### scales for ggplot2

```{r}
# color scale
scale_color_drsimonj <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- drsimonj_pal(palette = palette, reverse = reverse)
  
  if (discrete) {
    ggplot2::discrete_scale(
      aesthetics = "colour", 
      scale_name = paste0("drsimonj_", palette), 
      palette = pal,
      ...
    ) 
  } else {
    scale_colour_gradientn(colours = pal(256), ...)
  }
}

# fill scale
scale_fill_drsimonj <- function(palette = "main", discrete = TRUE, reverse = FALSE, ...) {
  pal <- drsimonj_pal(palette = palette, reverse = reverse)
  
  if (discrete) {
    ggplot2::discrete_scale(
      aesthetics = "fill",
      scale_name = paste0("disimonj_", palette),
      palette = pal,
      ...
    ) 
  } else {
      scale_fill_gradientn(colours = pal(256), ...)
  }
}


```


#### testing these palettes

```{r}
ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point(size = 4) + 
  scale_color_drsimonj()

ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Sepal.Length)) +
  geom_point(size = 4) + 
  scale_color_drsimonj(discrete = FALSE, palette = "cool")
```

```{r}
ggplot(mpg, aes(manufacturer, fill = manufacturer)) +
  geom_bar() +
  theme(
    axis.text = element_text(angle = 35, hjust = 1)
  ) +
  scale_fill_drsimonj(palette = "mixed", guide = "none")
```

