---
title: "stat summary of Ggplot2"
output:
  github_document:
    toc: true
    toc_depth: 3 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)

```


> **DISCLAIMER:** This note is based on (mostly copy pasted from) these sources
> [1] [demystifying-stat-layers-ggplot2](https://yjunechoe.github.io/posts/2020-09-26-demystifying-stat-layers-ggplot2/) [2] [Summary statistics from ggplot2tor](https://ggplot2tor.com/tutorials/summary_statistics)

## Demystifying stat_ layers in {ggplot2}

`stat_summary` works in the following order:

1. The data that is passed into `ggplot()` is inherited to `stat_summary` if one is not provided

2. The function passed into the `fun.data` argument applies transformations to a part of data (that was inherited/provided). `fun.data` defaults to `mean_se` function.

3. The result is then passed into `geom` provided in the `geom` argument of the `stat_summary` (`geom` defaults to `pointrange` if not specified).

4. If the transformed data contains all the required mappings for the geom, then geom will be printed.


- **`stat_summary` summarizes one dimension of the data.**


```{r understanding-stat-summary-1, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tibble)
```



```{r understanding-stat-summary-2}
height_df <- tibble(group = "A",
                    height = rnorm(30, 170, 10))

height_df %>% 
  ggplot(aes(x = group, y = height)) +
  stat_summary()
```


So as the points mentioned above, `geom = pointrange` and `fun.data = mean_se` are used here.

To verify this, we can actually look into the transformed data by `mean_se` and by `stat_summary`.

The `mean_se` function internally looks like,

```{r understanding-stat-summary-3}
mean_se

mean_se(height_df$height)
```

```{r understanding-stat-summary-4}
point_range_plot <- height_df %>% 
  ggplot(aes(group, height)) + 
  stat_summary()

layer_data(point_range_plot, 1) # here 1 means 1st layer which is pointrange here,

# but since here only one geom is used, so if we omit 1, consequence will be same.
```

which is similar as above, so it is proved that `stat_summary` using `mean_se` as default.

### Some Usecases of stat_summary

#### Error bar showing 95% confidence Interval


```{r understanding-stat-summary-5}
data(penguins, package = "palmerpenguins")

peng <- na.omit(penguins)
```


```{r understanding-stat-summary-6}
peng %>% 
  ggplot(aes(sex, body_mass_g)) +
  stat_summary(
    fun.data = ~mean_se(.x, mult = 1.96),
    geom = "errorbar"
  )
```


#### A color coded bar plot of median

Here we want to color a bar if median for a specific bar is less than a threshold (say 40).

```{r understanding-stat-summary-7}

# custom function for fun.data

calc_median_and_color <- function(x, threshold = 40) {
  tibble(y = median(x)) %>% 
    mutate(
      fill = if_else(y < threshold, "pink",  "gray35")
    )
}

peng %>% 
  ggplot(aes(species, bill_length_mm)) +
  stat_summary(
    fun.data = calc_median_and_color,
    geom = "bar"
  )

```


Here we have deliberately used the colname `fill` so that it can be used as an argument of geom.


#### pointrange plot with changing size

```{r understanding-stat-summary-8}
peng %>% 
  ggplot(aes(species, bill_length_mm)) +
  stat_summary(
    fun.data = \(x) {
      scaled_size <- length(x) / nrow(peng)
      mean_se(x) %>% mutate(size = scaled_size)
    }
  )
```



## Summary Statistic in Ggplot2 (tutorial from GGplot2tor)


```{r understanding-stat-summary-9, message=FALSE}
library(tidyverse)
library(gapminder)

head(gapminder)
```


```{r understanding-stat-summary-10}
gapminder %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_col()
```


here bar chart does not show the mean or median of lifeExpectancy for all countries, instead shows the sum of lifeExp for each year.

But if we want to show the mean lifeExp, we can do that.

```{r understanding-stat-summary-11}
gapminder %>% 
  ggplot(aes(x = year, y = lifeExp)) +
  geom_bar(fun = "mean", stat = "summary")

```



But in this approach, we can only draw bar plots. Heres where `stat_summary` comes into play. With `stat_summary` we can plot summary statistics with many different geoms.

```{r understanding-stat-summary-1-bis}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line")
```


If we want to see median lifeExps,

```{r understanding-stat-summary-2-bis}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(fun = "median", geom = "area", fill = "dodgerblue", alpha = 0.3) +
  stat_summary(fun = median, geom = "point", color = "firebrick")
```


Now if we want to plot point estimate with maximum minimum of lifeExp,


```{r understanding-stat-summary-3-bis}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(
    fun = mean,
    fun.min = min,
    fun.max = max,
    geom = "pointrange"
  )


# or using fun.data

gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(
    fun.data = \(x) {
      data.frame(y = mean(x), ymax = max(x), ymin = min(x))
    }
  )
```

we also can show error bar,


```{r understanding-stat-summary-4-bis}
gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(
    geom = "errorbar",
    width = 1,
    fun.min = min,
    fun.max = max
  )
```

```{r understanding-stat-summary-5-bis}
g <- gapminder %>% 
  ggplot(aes(year, lifeExp))

# 95% normal CI
g + stat_summary(
  fun.data = mean_cl_normal,
  geom = "errorbar",
  width = 1
) +
stat_summary(
  fun = mean,
  geom = "point"
)


# 99% normal CI
g + stat_summary(
  fun.data = ~ mean_cl_normal(.x, conf.int = 0.99),
  geom = "errorbar",
  width = 1
) +
stat_summary(
  fun = mean,
  geom = "point"
)
    
```


```{r understanding-stat-summary-6-bis}
median_IQR <- \(x) {
  list(
    y = median(x),
    ymin = quantile(x)[2],
    ymax = quantile(x)[4]
  )
}

range <- \(x) {
  list(ymin = min(x), ymax = max(x))
}
```


```{r understanding-stat-summary-7-bis}
g + 
  stat_summary(
    geom = "linerange",
    fun.data = median_IQR,
    size = 3,
    alpha = 0.3
  ) +
  stat_summary(
    geom = "linerange",
    fun.data = range,
    size = 3,
    alpha = 0.3
  ) +
  stat_summary(
    geom = "point",
    fun = "mean",
    shape = "X",
    size = 3
  )
```



```{r understanding-stat-summary-8-bis}
gapminder %>% 
  mutate(
    year = as.factor(year)
  ) %>% 
  ggplot(aes(continent, lifeExp, fill = year)) +
  stat_summary(
    geom = "bar",
    fun = "mean",
    position = position_dodge(0.95)
  ) +
  stat_summary(
    geom = "errorbar",
    fun.data = ~mean_cl_normal(.x, conf.int = 0.99),
    position = position_dodge(0.95),
    width = 0.5
  ) +
  stat_summary(
    geom = "point",
    fun = "mean",
    size = 1,
    position = position_dodge(0.95),
    show.legend = FALSE
  ) +
  scale_fill_viridis_d() +
  guides(
    fill = guide_legend(
      title = NULL,
      nrow = 2, byrow = TRUE, 
      keywidth = unit(20, "pt"),
      keyheight = unit(3, "pt")
    )
  ) +
  theme(
    legend.position = "bottom"
  )
```


```{r understanding-stat-summary-9-bis}
stat_fun <- \(x, fill = NULL, alpha = NULL, size = NULL) {
      list(
        y = mean(x), ymax = max(x), ymin = min(x),
        alpha = alpha, fill = fill, size = size
      )
}

gapminder %>% 
  ggplot(aes(year, lifeExp)) +
  stat_summary(
    geom = "ribbon",
    # fun.data = ~stat_fun(.x, fill = "firebrick", alpha = 0.3)
    fun.data = ~stat_fun(.x),
    fill = "firebrick",
    alpha = 0.3
  ) +
  stat_summary(
    geom = "pointrange",
    fun.data = ~stat_fun(.x),
    size = 0.4
  ) +
  stat_summary(
    geom = "line",
    fun = min,
    color = "firebrick"
  ) +
  stat_summary(
    geom = "line",
    fun = max,
    color = "firebrick"
  ) +
  facet_wrap(~continent)
```

